Do larger catchments respond different to forest cover change? Re-analysing a global data set.

### 2021-10-07 paper/Forest_and_Water/Forest_and_Water.Rmd
update paper 08102021

<!----
The starting point consists of the two tables in the supplementary data from Zhang et al. (2017), which means there is one table for small watersheds (< 1000 km2) and one table for large watersheds (> 1000 km2). We might combine these tables as we can easily add a variable indicating whether the watershed originates from the small or large table.
--->
```{r, echo = F, message = F}
require(tidyverse)
require(lubridate)

Zhang_small <- read_csv("data/TablesZhangetal2017_small.csv")
Zhang_large <- read_csv("data/TablesZhangetal2017_large.csv")
```

@@

However, the overall skewed distribution of the predictant ($\Delta \%Q$) is problematic, and this results in a skewed distribution of the GAM model residuals, which violates the linear model assumptions. As a result we transformed $\Delta \%Q$ and $\Delta \%forest$ back to fractions (0 - 1) and log transformed using $log10(x + 1)$, where $x$ is either $\Delta Q$ or $\Delta forest$. This means that the model residuals approximate $\sim N(0,\sigma^2)$ and this results in the following equation:

$$\tag{2}
\begin{aligned}
log10(\Delta Q) \sim ~ &log10(\Delta forest) + Pa + Area + Latitude + Longitude + \varepsilon
\end{aligned}$$

@@

$$\tag{3}
\begin{aligned}
log10(\Delta Q) \sim ~&log10(\Delta forest) + s(Pa, k = 3) + s(Area, k = 3) +  {forest~type} + \\  &{climate~type} + {assessment~type} + {hydrologic~regime} + \varepsilon
\end{aligned}$$

@@

$$\tag{4}
\begin{aligned}
log10(\Delta Q) \sim ~&log10(\Delta forest) + s(\frac{E0}{Pa}) + s(Area) +  {forest~type} + \\  &{climate~type} + {assessment~type} + {hydrologic~regime} + \varepsilon
\end{aligned}$$

@@

$$\tag{5}
\begin{aligned}
log10(\Delta Q) \sim ~&log10(\Delta forest) + s(Pa) + s(Area) + s(Latitude) + s(Longitude) + \\
& s(begin_{year}) + s(length_{study}) + {forest~type} + \\ & {climate~type} + {assessment~type} + {hydrologic~regime} + \varepsilon
\end{aligned}$$

$$\tag{6}
\begin{aligned}
log10(\Delta Q) \sim ~&log10(\Delta forest) + s(\frac{E0}{Pa}) + s(Area) + \\ 
& s(Latitude) + s(Longitude) + s(begin_{year}) + s(length_{study}) + \\ 
& {forest~type} + {climate~type} + {assessment~type} + \\  
& {hydrologic~regime} + \varepsilon
\end{aligned}$$

The results were analysed to identify:  
1. the significance of the different variables  
2. the direction of the categorical or shape of the smooth variables

@@

<!-- analysis -->
```{r model2, echo = F}
require(mgcv)
linear_model_all_log <- gam(log10(DeltaQf_perc/100+1) ~ DeltaF_perc + Area_km2 + Pa_mm, data = Zhang_all)
summary(linear_model_all_log)
gam.check(linear_model_all_log)
```
<!-- This shows that Area is not significant -->

@@

```{r model2a, echo = F}
Zhang_all <- Zhang_all %>%
  mutate(length = To - From)
linear_model_all_log <- gam(log10(DeltaQf_perc/100 + 1) ~ DeltaF_perc + log10(Area_km2) + Pa_mm +  Latitude + Longitude + From + length, data = Zhang_all)
summary(linear_model_all_log)
gam.check(linear_model_all_log)
```

@@

```{r}
Zhang_all %>%
  ggplot(aes(DeltaF_perc, log10(DeltaQf_perc/100 + 1), colour = Assessment_technique,size = log10(Area_km2))) + geom_point()
```
```{r}
tiff("flow_forest_byArea.tiff", width = 2000, height = 1400, res = 300)
Zhang_all %>%
  ggplot(aes(DeltaF_perc/100, (DeltaQf_perc/100 + 1), colour = Assessment_technique, size = log10(Area_km2))) + geom_point() +
  theme_bw() + ylab("log10 (fractional change in flow + 1)") +
  xlab("fractional change in forestry") + scale_y_log10() + scale_size_continuous(name = "log10 Area in km2") +
  scale_colour_discrete(name = "Assessment Technique")
dev.off()
```

@@

```{r}
Zhang_all %>%
  ggplot(aes(Area_km2)) + geom_histogram(fill="blue", bins =50) +
  scale_x_log10()
total <- nrow(Zhang_all)
length(Zhang_all$Area_km2[Zhang_all$Area_km2<10])/total
```

>>>>

<!----
The starting point consists of the two tables in the supplementary data from Zhang et al. (2017), which means there is one table for small watersheds (< 1000 km2) and one table for large watersheds (> 1000 km2). We might combine these tables as we can easily add a variable indicating whether the watershed originates from the small or large table.
--->
```{r, echo = F, message = F}
require(tidyverse)
require(lubridate)
require(mgcv)

Zhang_small <- read_csv("../../data/TablesZhangetal2017_small.csv")
Zhang_large <- read_csv("../../data/TablesZhangetal2017_large.csv")
```

@@

However, the overall skewed distribution of the predictant ($\Delta \%Q$) is problematic, and this results in a skewed distribution of the GAM model residuals, which violates the linear model assumptions. As a result we transformed $\Delta \%Q$ back to fractions (0 - 1) and log transformed using $log10(x + 1)$, where $x$ is $\Delta Q$.  After transformation the model residuals approximate $\sim N(0,\sigma^2)$ and this results in the following equation:

$$\tag{2}
\begin{aligned}
log10(\frac{\Delta \% Q}{100} + 1) \sim ~ &\Delta \%{forest~cover}
+ Pa + Area + Latitude + 
\\ &Longitude + \varepsilon
\end{aligned}$$

@@

$$\tag{3}
\begin{aligned}
log10(\frac{\Delta \% Q}{100} + 1) \sim ~ &\Delta \% forest~cover + s(Pa, k = 3) + s(Area, k = 3) +  \\
& {forest~type} + {climate~type} + {assessment~type} + \\  
& {hydrologic~regime} + \varepsilon
\end{aligned}$$

@@

$$\tag{4}
\begin{aligned}
log10(\frac{\Delta \% Q}{100} + 1) \sim ~&\Delta \% forest~cover + s(\frac{E0}{Pa}) + s(Area) +  {forest~type} + 
\\  &{climate~type} + {assessment~type} + \\  &{hydrologic~regime} + \varepsilon
\end{aligned}$$

@@

$$\tag{5}
\begin{aligned}
log10(\frac{\Delta \% Q}{100} + 1) \sim ~&\Delta \% forest~cover + s(Pa) + s(Area) + s(Latitude) + \\
& s(Longitude) + s(begin_{year}) + s(length_{study}) +\\
&{forest~type} + {climate~type} + {assessment~type} +\\
& {hydrologic~regime} + \varepsilon
\end{aligned}$$

$$\tag{6}
\begin{aligned}
log10(\frac{\Delta \% Q}{100} + 1) \sim ~&\Delta \% forest~cover +
s(\frac{E0}{Pa}) + s(Area) + \\ 
& s(Latitude) + s(Longitude) + s(begin_{year}) + \\ 
& s(length_{study}) + {forest~type} + {climate~type} + \\ 
& {assessment~type} + {hydrologic~regime} + \varepsilon
\end{aligned}$$
<!--
The results were analysed to identify:  
1. the significance of the different variables  
2. the direction of the categorical or shape of the smooth variables
-->

@@

<!-- analysis -->
Create a new data frame in which all the change in forestry is positive, but an additional column is added that indicates whether it was a forest cover increase or decrease.

```{r}
Zhang_all2 <- Zhang_all %>%
  mutate(Forest_Sign = ifelse(DeltaF_perc < 0,
                              "decrease", "increase"),
         DeltaF_perc_pos = ifelse(DeltaF_perc < 0,
                                  -1*DeltaF_perc,
                                  DeltaF_perc))

```

linear regression just looking at the change in forestry and the sign of the change (deforestation or forestation)

```{r model1, echo = F}
linear_model_simple <- gam(DeltaQf_perc ~ DeltaF_perc_pos +
                             Forest_Sign, 
                           data = Zhang_all2)
summary(linear_model_simple)
gam.check(linear_model_simple)

```
Change in forest cover has a significant impact on the change in flow, and the change in flow is different under deforestation compared to reforestation. In fact, reforestation has a 32% lower impact than deforestation. Overall model explanatory power is just `r summary(linear_model_simple)$r.squared`

```{r model2, echo = F}

linear_model2 <- gam(DeltaQf_perc ~ DeltaF_perc_pos + Forest_Sign + Area_km2 + Pa_mm, data = Zhang_all2)
summary(linear_model2)
gam.check(linear_model2)
```
Area is not a significant predictor of the change in flow.
But decreases in forestry definitely have a different effect from increases in flow.

<!-- This shows that Area is not significant -->

@@

```{r model2a1, echo = F}
Zhang_all <- Zhang_all %>%
  mutate(length = To - From)
linear_model_all_log <- gam(log10(DeltaQf_perc/100 + 1) ~ DeltaF_perc + log10(Area_km2) + Pa_mm +  Latitude + Longitude + From + length, data = Zhang_all)
summary(linear_model_all_log)
gam.check(linear_model_all_log)
```

@@

```{r}
Zhang_all2 %>%
  ggplot(aes(DeltaF_perc_pos, DeltaQf_perc, colour = Assessment_technique,size = log10(Area_km2), shape = Forest_Sign)) + geom_point()
```
```{r}
tiff("flow_forest_byArea.tiff", width = 2500, height = 1800, res = 300)
Zhang_all2 %>%
  ggplot(aes(DeltaF_perc_pos, DeltaQf_perc, colour = Assessment_technique,size = log10(Area_km2), 
             shape = Forest_Sign)) + geom_point(alpha = 0.5) +
  theme_bw() + ylab("% change in flow") +
  theme(axis.title = element_text(size = rel(2)),
        axis.text = element_text(size = rel(1.5))) +
  xlab("% change in forestry") + #scale_y_log10() +
  scale_size_continuous(name = "log10(Area in km2)") +
  scale_colour_discrete(name = "Assessment Technique") +
  scale_shape_discrete(name = "Forest cover direction")
dev.off()
```

@@

```{r}
Zhang_all %>%
  ggplot(aes(Area_km2)) + geom_histogram(fill="blue", bins =50) +
  scale_x_log10()
total <- nrow(Zhang_all)
length(Zhang_all$Area_km2[Zhang_all$Area_km2<10])/total
```
```{r}
Zhang_all %>%
  ggplot(aes(length)) + geom_histogram(fill="blue", bins =50) 
```
```{r}
#windows()
Zhang_all %>%
  ggplot(aes(log10(length), (DeltaQf_perc/100 + 1), colour = DeltaF_perc/100, size = log10(Area_km2))) + geom_point() +
  theme_bw() + ylab("log10 (fractional change in flow + 1)") +
  xlab("Length of study") + scale_size_continuous(name = "log10 Area in km2") +
  scale_colour_continuous(name = "change in forest cover")
```
